---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r echo=TRUE}
data=read.csv("activity.csv")
data$date = as.Date(data$date)

# adding necessary libraries for functions
library(dplyr)
library(ggplot2)
library(gridExtra)

```



## What is mean total number of steps taken per day?

#dataframe for total number of steps taken per day
```{r echo=TRUE}
steps_per_day <-data%>%select(date,steps)%>%group_by(date)%>%summarise(daily_steps=sum(steps,na.rm = TRUE))
colnames(steps_per_day) = c("date", "steps")
head(steps_per_day)
```

#histogram for the total number of steps taken per day
```{r plot_steps_per_day, fig.width=10, fig.height=4, echo=TRUE}

p = ggplot(steps_per_day, aes(x=date, y=steps)) + 
    geom_bar(stat="identity")+ scale_x_date(date_breaks="1 day" )
print(p)
```

#mean and median of number of steps taken per day
```{r echo=TRUE}
mean(steps_per_day$steps)
median(steps_per_day$steps)
```



## What is the average daily activity pattern?

#dataframe containing average number of steps taken per interval 
```{r echo=TRUE}
steps_per_interval <- data%>%select(interval,steps)%>%group_by(interval)%>%summarise(steps=mean(steps,na.rm = TRUE))
colnames(steps_per_interval) <- c("interval", "steps")
head(steps_per_interval)

p2<-ggplot(steps_per_interval, aes(interval, steps)) + geom_line() + theme_bw()
print(p2)

```

#finding the interval with the maximum number of steps
```{r echo=TRUE}
steps_per_interval[which.max(steps_per_interval$steps),]$interval
```



## Imputing missing values

#calculating total number of missing values

```{r echo=TRUE}
sum(is.na(data$steps))
```
#replacing missing value with the mean value observed for that interval, and creating new updated data set(data_imputed)

```{r echo=TRUE}
data_imputed = data
for (i in 1:nrow(data_imputed))
{
    row = data_imputed[i,]
    
    if (is.na(row$steps))
        data_imputed[i,"steps"] = steps_per_interval[which(steps_per_interval$interval == row$interval),1]
        
        # floor(steps_per_interval[steps_per_interval$interval==row$interval,]$steps)
}
```

#updated dataframe for total number of steps taken per day

```{r echo=TRUE}
steps_per_day_updated <-data_imputed%>%select(date,steps)%>%group_by(date)%>%summarise(daily_steps=sum(steps,na.rm = TRUE))
colnames(steps_per_day_updated) = c("date", "steps")
head(steps_per_day_updated)
```
#updated histogram for the total number of steps taken per day
```{r plot_steps_per_day, fig.width=10, fig.height=4, echo=TRUE}

p3 = ggplot(steps_per_day_updated, aes(x=date, y=steps)) + 
    geom_bar(stat="identity")+ scale_x_date(date_breaks="1 day" )
print(p3)

```
#mean and median of number of steps taken per day
```{r echo=TRUE}
mean(steps_per_day_updated$steps)
median(steps_per_day_updated$steps)
```
#these updated values for mean and median do differ from the original mean and median


## Are there differences in activity patterns between weekdays and weekends?

#update the imputed dataframe by adding a column indicating it as a weekday or not
```{r echo=TRUE}
data_imputed_days = data_imputed
data_imputed_days$daytype = ifelse(weekdays(data_imputed_days$date) %in% c("Saturday", "Sunday"),
                                       "weekend",
                                       "weekday")
data_imputed_days$daytype <- as.factor(data_imputed_days$daytype)



```

#create a plot containing average number of steps per interval on weekdays and weekends
```{r plot_data_weekday_steps_per_interval, fig.width=10, fig.height=6, echo=TRUE}
# plot for weekends
data_weekend = data_imputed_days[data_imputed_days$daytype == "weekend", ]
data_weekend_steps_per_interval = aggregate(data_weekend$steps, 
                                  by=list(interval=data_weekend$interval), 
                                  FUN=mean)
colnames(data_weekend_steps_per_interval) = c("interval", "steps")
p4 = ggplot(data_weekend_steps_per_interval, aes(interval, steps)) + 
    geom_line() + theme_bw()  + ggtitle("weekends")
# plot for weekdays
data_weekday = data_imputed_days[data_imputed_days$daytype == "weekday", ]
data_weekday_steps_per_interval = aggregate(data_weekday$steps, 
                                    by=list(interval=data_weekday$interval), 
                                    FUN=mean)
colnames(data_weekday_steps_per_interval) = c("interval", "steps")
p5 = ggplot(data_weekday_steps_per_interval, aes(interval, steps)) + 
    geom_line() + theme_bw()  + ggtitle("weekdays")
# stack plots
grid.arrange(p5, p4, nrow=2)
```

